Assessing the Population-Level Conservation Effects of Marine Protected Areas

### 2020-12-01 documents/ovando-et-al-pop-effects-of-mpas.Rmd
Edits almost done

---
title: Assessing the Population-Level Conservation Effects of Marine Protected Areas
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world’s ocean, up from less than 1% in the year 2000, and international commitments call for 10%, 30%, and even 50% coverage. The premise underlying MPA expansion is that they conserve biodiversity, habitats, and fished populations. While numerous studies show that MPAs produce conservation benefits *inside* their borders, many MPAs are also justified on the grounds that they confer conservation benefits to the broader population *beyond* their borders. We examine the conditions under which MPAs can provide population-level conservation benefits inside and outside their borders, and show that even in cases where the population benefits are large, they are inherently difficult to detect empirically. A network of MPAs was put in place in The Channel Islands National Marine Sanctuary in 2003, with a goal of providing regional conservation and fishery benefits. Evidence indicates that the Channel Island MPAs have increased biomass densities inside the MPAs, but we are unable to find a clear effect of these same MPAs at the population level. Using a Bayesian difference-in-difference approach, we show that MPA effect sizes less than 30% are likely to be difficult to detect (even when they are present); the size of many MPA networks suggests that effect sizes may often be smaller than 30%. Our results provide a novel assessment of the population-level effects of a large and iconic Marine Protected Area network, and provide guidance for communities charged with monitoring and adapting MPAs.
bibliography: references.yaml
csl: conservation-biology.csl
output:
  bookdown::word_document2:
    reference_docx: template.docx
  bookdown::pdf_document2:
    keep_tex: true
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
  - \usepackage{draftwatermark}
  - \SetWatermarkText{DRAFT}
always_allow_html: true
---

@@

No-take Marine Protected Areas (MPAs), spatial regions of the ocean in which fishing is prohibited, have a long history in the management of marine resources. Modern MPAs were first established as marine analogs to the terrestrial protection of iconic landscapes [@iucn1976]. Over time our goals and expectations for MPAs have evolved and become more ambitious. Recent international efforts to expand MPAs, such as the Biodiversity Beyond National Jurisdiction (BBNJ) Agreement and the Convention on Biological Diversity’s Post-2020 Global Biodiversity Framework, are based in part on the assumption that well-designed MPAs will not only provide conservation benefits inside their borders, but also have broader conservation effects on unprotected areas surrounding the MPAs [@gaines2010].

The empirical MPA literature has focused on assessing the ability of MPAs to provide conservation gains within their borders [@lester2009; @edgar2014; @gerber2005]. However, as conservation benefits accrue inside MPAs, MPAs also affect the waters beyond their borders through the spillover of adult and larval fish from the protected to the fished areas, as well as through displacement of fishing effort. Therefore, MPAs contribute to both local and population-level effects (i.e., inside-MPA effects and broader regional effects). Numerous factors influence how MPAs affect fish populations. These include the scale of adult and larval dispersal relative to the size of the MPAs [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008;@gerber2005], the strength, timing, and location of density dependence [@burgess2014], the design of the network [@gaines2010; @costello2010a],the degree of enforcement [@edgar2014], the level of fishing pressure, the time span under evaluation, and how fishing and management responds to the implementation of the MPAs [@hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@smith2006a;@walters2000a;@white2011;@white2010c;@ovando2016;@moffitt2013].

<!-- What empirical evidence is there for these population-level effects of MPAs?  -->
These theoretical results are generally based on modeling of closed populations, where some fraction of a population is protected inside of MPAs. In contrast to this population paradigm used in MPA simulations, MPAs are often evaluated empirically at local scales using spatial "response ratios", commonly measured as the ratio of biomass densities (weight of organisms per unit area) of species inside relative to selected control sites outside of MPAs [@lester2009; @halpern2003a;@edgar2014;@caselle2015]. These studies found clear evidence that well enforced and sufficiently sized MPAs are associated with high response ratios. Several studies have also documented empirical evidence for the existence of adult or larval fish spillover affecting fish abundance [@halpern2009;@kay2012; @stobart2009; @mcclanahan2000; @russ1996; @thompson2017;@baetscher2019;@pelc2009]. Where response ratios are available before and after MPA implementation, spatial before-after-control-impact (BACI) style studies have shown similarly clear and positive results [@thiault2019a].

While these studies have clearly demonstrated the ability of MPAs to create differences between local fished and unfished areas, this body of empirical work does not necessarily serve as sufficient evidence for the population-level effects of MPAs. This is because these studies rely on assuming that selected control sites serve as a measure of what would have happened in the absence of MPAs. Control sites used in calculating response ratios are often selected based on habitat characteristics to justify their use as counterfactuals [@ferraro2018]. However, the very spillover effects MPAs are increasingly hoped to produce can negate the ability of spatial response ratio / BACI designs to accurately estimate the effects of MPAs. Export of adults or larvae from MPAs to "control" sites can affect their status as controls, as does displacement of fishing effort from MPAs to control sites. These factors can bias estimators of population-level effects that require the assumption that the control sites are unaffected by the treatment (a required assumption of response ratios and BACI designs [@ferraro2018;@moffitt2013]). Control sites sufficiently far from MPAs to negate both spillover of fish or larvae and concentration of the fishing fleet could be selected, but finding suitably distant sites that are also appropriate proxies for the ecological and economic context of the MPAs is challenging. As variations of spatial response ratios and BACI studies are our primary source of evidence for the conservation effects of MPAs, this means that our empirical understanding of the population-level impacts of MPAs is surprisingly limited.

@@

We present here an paired theoretical and empirical assessment demonstrating the challenges in assessing the population-level impacts of MPAs. In 2003, a network of MPAs was established in the Channel Islands National Marine Sanctuary, California, USA (hereafter the Channel Islands) covering approximately 20% of the Islands' waters [@kirlin2013]. The network has been used as a model case study in protected area design around the world (@botsford2014 and references therein). We use data from the first 14 years of protection in a difference-in-difference model [see @angrist2009 and references therein] to provide what is to our knowledge the first empirical assessment of the population-level effect of a large MPA network on a wide array of fin-fish species. Rather than relying on spatial controls as is typically done in MPA studies, we use groups of species targeted and not-targeted by fishing pressure as our treatment and control groups. In contrast to clear differences in biomass densities observed inside and outside of well-protected MPAs both globally [@lester2009] and in the Channel Islands [@caselle2015] we are unable to detect a clear population effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the population-level conservation effects of MPAs.

@@

Our primary methods consist of a spatially-explicit bio-economic simulation model and a Bayesian difference-in-difference regression [@angrist2009;@mcelreath2020]. Difference-in-difference is a regression-based approach to BACI, assessing changes in control and treatment groups before and after treatment [@larsen2018]. We use the term difference-in-difference here to more closely tie our results to the terminology in causal-inference literature. The population-level effects of MPAs will depend on the social-ecological contexts in which they occur. We use a bio-economic simulation model to provide theoretical bounds on potential population-level effects of MPAs, which we then confront with the empirical results from our difference-in-difference model.

All analysis were conducted in R [@rcoreteam2019]. Our difference-in-difference model was fit using Stan [@carpenter2017] using the `rstanarm` package [@goodrich2020]. All data and code needed to fully reproduce this manuscript are publicly available at [github.com/DanOvando/population-effects-of-mpas](https://github.com/DanOvando/population-effects-of-mpas). A detailed description of the simulation model structure, as well as sensitivity analyses of our estimation model, are available in the online supporting information (SI).

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a fish population. Readers can explore the functionality of the model using an online tool available at [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/). The purpose of the simulation model is to both set expectations for our empirical results, and demonstrate more broadly the ways in which ecological and economic dynamics can interact to produce a wide range of population-level MPA effects. The full range of factors explored are shown in  (Table.\@ref(tab:sim-table)), and the equations of the simulation model can be viewed in the SI section XX. We used this model to generate `r n_distinct(outcomes$experiment)` simulated MPA outcomes across `n_distinct(outcomes$scientific_name)` species.

Many papers have presented simulation analyses of MPA outcomes [@fulton2015]. Our model incorporates a broad range of core ecological and economic drivers of MPA performance assessed by individual papers into a cohesive model, similar in spirit to @krueck2017. The simulation model consists of 50 patches with wrapped edges. For each simulation we first randomly pull a species and its associated life history traits from the `FishLife` [@thorson2017c] package. We pair these data with randomly selected values governing the characteristics of the simulation (Table.\@ref(tab:sim-table)). Key choices available to the model include parameters governing fishing pressure and MPA design. For a given simulation, the model randomly selects a fleet model and fishing effort allocation strategy. The fleet model can either be constant-catch (the fleet exerts as much effort as needed to maintain a fixed amount of catch), constant-effort (the fleet maintains a constant amount of effort over time), or open-access (fishing effort of the fleet expands and contracts in response to available profits). The total fishing effort exerted by the fleet can then be distributed in space uniformly, in proportion to catch per unit effort, or in proportion to profit per unit effort.

The simulation then begins to apply the fleet model to the population, and in a randomly selected year implements an MPA network. The model samples a percentage of the population's range to place in MPAs, and randomly assigns patches to MPAs either across across a uniform system or preferentially on higher-quality habitat. The model then randomly selects whether fishing effort that used to operate inside the MPAs is redistributed to the areas outside the MPAs, or leaves the fishery entirely. We then continue the simulations with the MPAs in place. Each simulation is paired with a simulation identical in every way except that MPAs are not implemented (i.e., a simulated control). We then calculate the effect of the MPAs on the population as the difference in relevant metrics in the simulation with MPAs relative to their values in the simulation without MPAs.

These simulation results provide a library of plausible MPA effects for a range of biological and economic assumptions. One set of simulations is specifically designed to reflect the dynamics of the Channel Islands. We only include species of the same genus as those targeted by fishing in the Channel Islands, restrict fishing pressure such that the simulated populations are moderately to lightly exploited (given the best available knowledge of stock status for the targeted species in the Channel Islands), and cap the MPA size at 20% of the population's range [@rassweiler2012]. For these Channel Island-esque simulations, we calculated 1) the population-level response for each species (the 'true' population-wide difference in biomass between the simulations with and without the MPAs) and 2) the measured response ratio of biomass densities inside and outside the simulated MPAs. To compare response ratios from the Channel Islands simulations to our empirical results, we organized simulated and empirical response ratios into bins. Because each simulation included measures of both response ratios and population-level effects, this provided a library of simulations (and their associated attributes) that could have produced the types of empirical response ratios measured in the Channel Islands.

@@

The difference-in-difference model used empirical kelp forest survey data from the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) monitoring in the Northern Channel Islands. PISCO conducts visual SCUBA surveys at a large number of rocky reef and kelp forest sites inside and outside of MPAs throughout the Channel Islands, producing estimates of densities of fishes that are both targeted and non-targeted by fishing (Fig.\ref{fig:ci-map},Fig.\@ref{fig:raw-density}). The details of the monitoring program are described in @caselle2015. We define the population-level conservation effects of MPAs as the change in mean total biomass densities of targeted fin-fish both inside and outside of MPAs, relative to the mean total biomass densities of targeted fin-fish inside and outside of MPAs that would have occurred without the MPAs.

Building off of the concepts explored in @caselle2015, we used an identification strategy utilizing biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 0])` species that are not directly targeted by fishing as our control group (non-targeted), and biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 1])` species targeted by fishing as our treatment group (see Fig.\@ref(fig:raw-density) for raw trends in targeted and non-targeted biomass densities). Targeted species in the Channel Islands include fin-fish such as California sheephead (*Semicossyphus pulcher*), and copper (*Sebastes caurinus*) and blue (*Sebastes mystinus*) rockfish. Non-targeted species include garibaldi (*Hypsypops rubicundus*), halfmoons (*Medialuna californiensis*), and blacksmith (*Chromis punctipinnis*). We used a Bayesian difference-in-difference regression to estimate any difference in mean total biomass densities of fin-fish species targeted by fishing effort (i.e., those potentially affected by an MPA) and those species not targeted by fishing before and after MPA implementation. Note that this is akin to a BACI, but rather than relying on spatial control sites as our identification, we utilize non-targeted species as our control. To account for the fact that sampling locations are not uniformly distributed across the islands, we weight the samples in our regression in proportion to the area inside and outside of MPAs in the Channel Islands.

This identification strategy controls for unobserved environmental shocks to the system that are independent of the MPAs. Conditional on the assumptions of the model, this regression produces an estimate of the effect of the MPAs on the mean total biomass densities of targeted species throughout the Channel Islands. For example, consider an evenly distributed population that has 50% of its range protected by an MPA. If the MPA increases biomass densities inside its boundaries by 20%, and by 0% outside the reserve, the population effect of the MPA would be 10%.

The difference-in-difference (DiD) regression amounts to estimating the pre-post MPA difference in the biomass densities of targeted species, minus the same difference for non-targeted species in the Channel Islands.

@@

where $d_i$ is the biomass density at observation *i*, *T* indicates whether the observation *i* is for a targeted ($T = 1$) or non-targeted ($T = 0$) species, and *MPA* marks whether observation *i* is in a pre MPA ($MPA = 0$) or post MPA ($MPA = 1$) state. **B^c^** is a vector of coefficients for additional control variables in matrix *X* such as water visibility and observer experience. **B^s^** is a vector of hierarchical coefficients for each sampling location **S** , clustered by island.  Under the assumptions of this model, $\beta_3$ is the causal effect of the treatment (*MPA*) on the treated (targeted species). $\beta_0$ is the mean total biomass of non-targeted species pre-MPA, and $\beta_1T$ is the mean total biomass of targeted species pre-MPA. $\beta_2MPA$ is the effect of the post-MPA period on the non-targeted species. The shape and scale parameters of the Gamma distribution are estimated as well (Table.\@ref(tab:coef-table)).

This difference-in-difference regression provides our estimate of the population-level effect of the Channel Islands MPAs. We use a Bayesian regression since it allows us to interpret our estimated effects probabilistically [@mcelreath2020], e.g. stating the probability that conditional on the model, priors,and data the Channel Island MPAs increased or decreased biomass of targeted species. Being a Bayesian regression, our difference-in-difference model produces posterior probability distributions of our coefficients, from which we can construct Bayesian credible intervals. Since the effects of MPAs will evolve over time, we use this model to estimate a series of population-level effects of the Channel Islands MPAs in three-year intervals from 2003 to 2017.

@@

Our Channel Islands-specific simulation provides a set of outcomes that can feasibly bound our expectations for the empirical effects of the MPA network. The targeted species in our database span a range of life histories, but are largely made up of fishes in the perch and rockfish complexes, with a mean Von Bertalanffy growth coefficient of `r round(mean(used_data$k[used_data$targeted == 1]),2)` (standard deviation `r round(sd(used_data$k[used_data$targeted == 1]),2)`), and a median age at maturity of `r round(median(used_data$tm[used_data$targeted == 1]),0)` years (standard deviation `r round(sd(used_data$tm[used_data$targeted == 1]),0)`), and a mix of exploitation histories, ranging from low to high (but mostly relatively low) [@alonzo2004;@dick2017]. While a wide range of MPA effects are plausible, from 0% to upwards of 150%, the median simulated effect size for this stylized version of the Channel Islands after over ten years of protection was `r percent(median(sim_mpa_effects$mpa_effect[sim_mpa_effects$years_protected > 10]))`.

We compared this range of plausible outcomes to the empirical data. @caselle2015 found a statistically significant increase in the response ratios of targeted species over time, and evidence that this increase is smaller for non-targeted species. Updating the results of @caselle2015 through 2017 shows a continuation of the increasing trend in the response ratios of targeted species (Fig.\@ref(fig:rr-plot)). This provides evidence that the Channel Islands MPAs are large enough and sufficiently well-enforced as to provide meaningful protection within their borders.

These response ratios cannot, however, be used as a definitive indicator of population effects of the MPAs. In the case of the Channel Islands MPAs, control sites are often located within a few kilometers of an MPA, suggesting that they are susceptible to both local biological spillover and by concentration of fishing effort excluded from the MPAs. Our simulation results show that the response ratio trends we observe in the data could plausibly be produced by a wide range of population-level MPA effects (Fig.\@ref(fig:rr-plot)). Response ratios well over one were associated with "true" population-level MPA effects generally less than 25%, and, importantly, many simulations produced large response ratios but population-level MPA effects close to 0%. This can occur if for example fishing pressure is only moderate, adult movement is low, larval dispersal is high, and displaced fishing effort concentrates around the border of the MPAs.

@@

Containing a carefully designed, well-enforced, and well-studied MPA network, the Channel Islands seems to be an ideal location to study the population-level effects of protected areas. Persistently high response ratios suggest that despite overall decreases in targeted biomass densities inside and outside MPAs relative to pre-MPA measurements (Fig.\@ref(fig:raw-density)), the MPAs may still be providing protection within their borders. But, these response ratios are not necessarily an indicator of broader population-level effects. The targeted vs. non-targeted difference-in-difference strategy presents an alternative means of estimating the population-level effects of MPAs. While having its own strict caveats, this strategy provides some potential improvements over spatial response-ratios as a means of estimating population-level effects. While we estimate an uncertain but overall positive effect of the MPA network in its first few years, we are unable to detect a robust positive signal from 2012-2017. After 14 years of MPA protection we are left without a clear picture of the population-level effect of the Channel Island MPA network on biomass densities of targeted fin-fish species.

@@

@@

@@

knitr::kable(equations,
             escape = FALSE,
             caption = "Variables for MPA simulations",
             format = "pipe") %>% 
  kableExtra::kable_styling()

@@

```{r coef-table, results = "asis", echo=FALSE}

>>>>

---
title: Assessing the Population-Level Conservation Effects of Marine Protected Areas
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world’s ocean, up from less than 1% in the year 2000, and international organizations call for 30% by 2030. The premise underlying MPA expansion is that they conserve biodiversity, habitats, and fished populations. While numerous studies show that MPAs produce conservation benefits *inside* their borders, many MPAs are also justified on the grounds that they confer conservation benefits to the broader population *beyond* their borders. We examine the conditions under which MPAs can provide population-level conservation benefits inside and outside their borders, and show that even in cases where the population benefits are large, they are inherently difficult to detect empirically. A network of MPAs was put in place in The Channel Islands National Marine Sanctuary in 2003, with a goal of providing regional conservation and fishery benefits. Evidence indicates that the Channel Island MPAs have increased biomass densities inside the MPAs, but we are unable to find a clear effect of these same MPAs at the population level using a Bayesian difference-in-difference appreoach. We show that MPA effect sizes less than 30% are likely to be difficult to detect (even when they are present); the size of many MPA networks suggests that effect sizes may often be smaller than 30%. Our results provide a novel assessment of the population-level effects of a large and iconic Marine Protected Area network, and provide guidance for communities charged with monitoring and adapting MPAs.
bibliography: references.yaml
csl: conservation-biology.csl
output:
  bookdown::word_document2:
    reference_docx: template.docx
  bookdown::pdf_document2:
    keep_tex: true
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
  - \usepackage{draftwatermark}
  - \SetWatermarkText{DRAFT}
always_allow_html: true
---

@@

No-take Marine Protected Areas (MPAs), spatial regions of the ocean in which fishing is prohibited, have a long history in the management of marine resources. Modern MPAs were first established as marine analogs to the terrestrial protection of iconic landscapes [@iucn1976]. Over time our goals and expectations for MPAs have evolved and become more ambitious. Recent international efforts to expand MPAs, such as The International Union for Conservation of Nature's 30% by 2030 MPA targets are based in part on the assumption that well-designed MPAs will not only provide conservation benefits inside their borders, but also have broader conservation effects on unprotected areas surrounding the MPAs [@gaines2010].

The empirical MPA literature has focused on assessing the ability of MPAs to provide conservation gains within their borders [@lester2009; @edgar2014]. However, as conservation benefits accrue inside MPAs, MPAs also affect the waters beyond their borders through the spillover of adult and larval fish from the protected to the fished areas, as well as through displacement of fishing effort. Therefore, MPAs contribute to both local and population-level effects (i.e., inside-MPA effects and broader regional effects). Numerous factors influence how MPAs affect fish populations. These include the scale of adult and larval dispersal relative to the size of the MPAs [@gaines2003; @botsford2008], the strength, timing, and location of density dependence [@burgess2014], the design of the network [@gaines2010; @costello2010a],the degree of enforcement [@edgar2014], the level of fishing pressure, the time span under evaluation, and how fishing and management responds to the implementation of the MPAs [@hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@walters2000a;@white2011;@ovando2016;@moffitt2013].

<!-- What empirical evidence is there for these population-level effects of MPAs?  -->
These largely theoretical results are generally based on modeling of closed populations with some fraction protected inside of MPAs. In contrast to this population paradigm used in MPA simulations, MPAs are often evaluated empirically at local scales using spatial "response ratios", commonly measured as the ratio of biomass densities (weight of organisms per unit area) of species inside relative to selected control sites outside of MPAs [@lester2009; @halpern2003a;@edgar2014;@caselle2015]. These studies found clear evidence that well enforced and sufficiently sized MPAs are associated with high response ratios. Several studies have also documented empirical evidence for the existence of adult or larval fish spillover affecting fish abundance [e.g. @halpern2009;@kay2012; @mcclanahan2000; @russ1996]. Where response ratios are available before and after MPA implementation, spatial before-after-control-impact (BACI) style studies have shown similarly clear and positive results [@thiault2019a].

While these studies have clearly demonstrated the ability of MPAs to create differences between local fished and unfished areas, this body of empirical work does not necessarily serve as sufficient evidence for the population-level effects of MPAs. These spatial studies rely on assuming that selected control sites serve as a measure of what would have happened in the absence of MPAs. Control sites used in calculating response ratios are often selected based on habitat characteristics to justify their use as counterfactuals [@ferraro2018]. However, beyond habitat differences, the very spillover effects MPAs are increasingly hoped to produce can negate the ability of spatial response ratio or BACI designs to accurately estimate the effects of MPAs. Export of adults or larvae from MPAs to "control" sites can affect their status as controls, as does displacement of fishing effort from MPAs to control sites. These factors can bias estimators of population-level effects like response ratios that require the assumption that the control sites are unaffected by the treatment [@ferraro2018;@moffitt2013]. Control sites sufficiently far from MPAs to negate both spillover of fish or larvae and concentration of the fishing fleet could be selected, but finding suitably distant sites that are also appropriate proxies for the ecological and economic context of the MPAs is challenging. As variations of spatial response ratios and BACI studies are our primary source of evidence for the conservation effects of MPAs, this means that our empirical understanding of the population-level impacts of MPAs is surprisingly limited.

@@

We present here an paired theoretical and empirical assessment demonstrating the challenges in assessing the population-level impacts of MPAs. In 2003, a network of MPAs was established in the Channel Islands National Marine Sanctuary, California, USA (hereafter the Channel Islands) covering approximately 20% of the Islands' waters. The network has been used as a model case study in protected area design around the world [@botsford2014 and references therein]. We use data from the first 14 years of protection in a difference-in-difference (DiD) model [see @angrist2009 and references therein] to provide what is to our knowledge the first empirical assessment of the population-level effect of a large MPA network on a wide array of fin-fish species. Rather than relying on spatial controls as is typically done in MPA studies, we use groups of species targeted and not-targeted by fishing pressure as our treatment and control groups. In contrast to clear differences in biomass densities observed inside and outside of well-protected MPAs both globally [@lester2009] and in the Channel Islands [@caselle2015] we are unable to detect a clear population effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the population-level conservation effects of MPAs.

@@

Our methods consist of a spatially-explicit bio-economic simulation model and a Bayesian DiD regression [@angrist2009;@mcelreath2020]. DiD is a regression-based approach to BACI, assessing changes in control and treatment groups before and after treatment [@larsen2018]. We use the term DiD here to more closely tie our results to the terminology in causal-inference literature. We used a bio-economic simulation model to provide theoretical bounds on potential population-level effects of MPAs, which we then confront with the empirical results from our DiD model.

All analysis were conducted in R [@rcoreteam2019]. Our DiD model was fit using Stan [@carpenter2017] using the rstanarm package [@goodrich2020]. All data and code needed to fully reproduce this manuscript are publicly available at [github.com/DanOvando/population-effects-of-mpas](https://github.com/DanOvando/population-effects-of-mpas). A detailed description of the simulation model structure, as well as sensitivity analyses of our estimation model, are available in the online Appendix Supporting Information 1.

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a fish population. Readers can explore the functionality of the model using an online tool available at [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/). The purpose of the simulation model is to both set expectations for our empirical results, and demonstrate the ways in which ecological and economic dynamics can interact to produce a wide range of population-level MPA effects. The full range of factors explored are shown in  Table.\@ref(tab:sim-table), and the equations of the simulation model can be viewed in the Appendix S1 Section 2. We used this model to generate `r n_distinct(outcomes$experiment)` simulated MPA outcomes across `r n_distinct(outcomes$scientific_name)` species.

Many papers have presented simulation analyses of MPA outcomes [@fulton2015]. Our model incorporates a broad range of core ecological and economic drivers of MPA performance assessed by these individual papers into a cohesive model, similar in spirit to @krueck2017. The simulation model consists of 50 patches with wrapped edges. For each simulation we first randomly pull a species and its associated life history traits from the `FishLife` [@thorson2017c] package. We pair these data with randomly selected values governing the characteristics of the simulation (Table.\@ref(tab:sim-table)). Key choices available to the model include parameters governing fishing pressure and MPA design. For a given simulation, the model randomly selects a fleet model and fishing effort allocation strategy. The fleet model can either be constant-catch (the fleet exerts as much effort as needed to maintain a fixed amount of catch), constant-effort (the fleet maintains a constant amount of effort over time), or open-access (fishing effort of the fleet expands and contracts in response to available profits). The total fishing effort exerted by the fleet can then be distributed in space uniformly, in proportion to spatial catch per unit effort, or in proportion to spatial profit per unit effort.

The simulation then begins to apply the fleet model to the population, and in a randomly selected year implements an MPA network. The model samples a percentage of the population's range to place in MPAs, and randomly assigns patches to MPAs either across across a uniform system or preferentially on higher-quality habitat. The model then randomly selects whether fishing effort that used to operate inside the MPAs is redistributed to the areas outside the MPAs, or leaves the fishery entirely. We then continue the simulations with the MPAs in place. Each simulation is paired with a simulation identical in every way except that MPAs are not implemented (i.e., a simulated control). We then calculate the effect of the MPAs on the population as the difference in biomass densities in the simulation with MPAs relative to biomass densities in the simulation without MPAs.

These simulation results provide a library of plausible MPA effects for a range of biological and economic assumptions. One set of simulations is specifically designed to reflect the dynamics of the Channel Islands. For this set we only include species of the same genus as those targeted by fishing in the Channel Islands, restrict fishing pressure such that the simulated populations are moderately to lightly exploited, and cap the MPA size at 20% of the population's range [@rassweiler2012]. For each of these Channel Island style simulations, we calculated 1) the 'true' population-wide difference in biomass between the simulations with and without the MPAs and 2) the response ratio of biomass densities inside and outside the simulated MPAs. We then calculated the empirical response ratios observed in the real Channel Islands, and matched these empirical results with simulations that produced similar response ratios after the same number of years of MPA protection. Because each simulation included measures of both response ratios and population-level effects, this process provides a library of simulations (and their associated attributes) that could have produced the types of empirical response ratios measured in the Channel Islands.

@@

The DiD model used empirical kelp forest survey data from the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) monitoring in the Northern Channel Islands. PISCO conducts visual SCUBA surveys at a large number of rocky reef and kelp forest sites inside and outside of MPAs throughout the Channel Islands, producing estimates of densities of fishes that are both targeted and non-targeted by fishing (Fig.\ref{fig:ci-map},Fig.\@ref{fig:raw-density}). The details of the monitoring program are described in @caselle2015. We define the population-level conservation effects of MPAs as the change in mean total biomass densities of targeted fin-fish both inside and outside of MPAs, relative to the mean total biomass densities of targeted fin-fish inside and outside of MPAs that would have occurred without the MPAs.

Building off of the concepts explored in @caselle2015, we used an identification strategy utilizing biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 0])` species that are not directly targeted by fishing as our control group (non-targeted), and biomass densities of `r n_distinct(used_data$classcode[used_data$targeted == 1])` species targeted by fishing as our treatment group (see Fig.\@ref(fig:raw-density) for raw trends in targeted and non-targeted biomass densities). Targeted species in the Channel Islands include fin-fish such as California sheephead (*Semicossyphus pulcher*), and copper (*Sebastes caurinus*) and blue (*Sebastes mystinus*) rockfish. Non-targeted species include garibaldi (*Hypsypops rubicundus*), halfmoons (*Medialuna californiensis*), and blacksmith (*Chromis punctipinnis*). Our regression estimates any difference in mean total biomass densities of fin-fish species targeted by fishing effort (i.e., those potentially affected by an MPA) and those species not targeted by fishing before and after MPA implementation. To account for the fact that sampling locations are not uniformly distributed across the islands, we weight the samples in our regression in proportion to the total area inside and outside of MPAs in the Channel Islands.

This identification strategy attempts to control for unobserved environmental shocks to the system that are independent of the MPAs. Conditional on the assumptions of the model, this regression produces an estimate of the effect of the MPAs on the mean total biomass densities of targeted species throughout the Channel Islands. For example, consider an evenly distributed population that has 50% of its range protected by an MPA. If the MPA increases biomass densities inside its boundaries by 20%, and by 0% outside the reserve, the population effect of the MPA estimated by our DiD would be 10%.

The DiD regression amounts to estimating the pre-post MPA difference in the biomass densities of targeted species, minus the same difference for non-targeted species in the Channel Islands.

@@

where $d_i$ is the biomass density at observation *i*, *T* indicates whether the observation *i* is for a targeted ($T = 1$) or non-targeted ($T = 0$) species, and *MPA* marks whether observation *i* is in a pre MPA ($MPA = 0$) or post MPA ($MPA = 1$) state. To account for the fact that MPA effects will evolve over time, we estimate a vector of MPA effects in three-year blocks for the all years after the MPAs were implemented in 2003. **B^c^** is a vector of coefficients for additional control variables in matrix *X* such as water temperature and observer experience. **B^s^** is a vector of hierarchical coefficients for each sampling location **S** , clustered by island.  Under the assumptions of this model, $\beta_3$ is the causal effect of the treatment (*MPA*) on the treated (targeted species). $\beta_0$ is the mean total biomass of non-targeted species pre-MPA, and $\beta_1T$ is the mean total biomass of targeted species pre-MPA. $\beta_2MPA$ is the effect of the post-MPA period on the non-targeted species (Table.\@ref(tab:coef-table)). We use a Bayesian regression since it allows us to interpret our estimated effects probabilistically [@mcelreath2020]. Being a Bayesian regression, our DiD model produces posterior probability distributions (the probability distribution of our coefficients conditional on the data, priors, and model assumptions) of our coefficients, from which we can construct Bayesian credible intervals [@gelman2014a].

@@

<!-- Our Channel Islands-specific simulation provides a set of outcomes that can feasibly bound our expectations for the empirical effects of the MPA network. The targeted species in our database span a range of life histories, but are largely made up of fishes in the perch and rockfish complexes, with a mean Von Bertalanffy growth coefficient of `r round(mean(used_data$k[used_data$targeted == 1]),2)` (standard deviation `r round(sd(used_data$k[used_data$targeted == 1]),2)`), and a median age at maturity of `r round(median(used_data$tm[used_data$targeted == 1]),0)` years (standard deviation `r round(sd(used_data$tm[used_data$targeted == 1]),0)`), and a mix of exploitation histories, ranging from low to high (but mostly relatively low) [@alonzo2004;@dick2017].  -->

@caselle2015 found a statistically significant increase in the response ratios of targeted species over time, and evidence that this increase is smaller for non-targeted species. Updating the results of @caselle2015 through 2017 shows a continuation of the increasing trend in the response ratios of targeted species (Fig.\@ref(fig:rr-plot)). This provides evidence that the Channel Islands MPAs are large enough and sufficiently well-enforced as to provide meaningful protection within their borders.

These response ratios cannot, however, be used as a definitive indicator of population effects of these MPAs. In the case of the Channel Islands MPAs, control sites are often located within a few kilometers of an MPA, suggesting that they are susceptible to both local biological spillover and by concentration of fishing effort excluded from the MPAs. Our simulation results show that the response ratio trends we observe in the data could plausibly be produced by a wide range of population-level MPA effects (Fig.\@ref(fig:rr-plot)). Response ratios well over one were associated with "true" population-level MPA effects generally less than 25%, and, importantly, many simulations produced large response ratios but population-level MPA effects close to 0%. This can occur if for example fishing pressure is only moderate, adult movement is low, larval dispersal is high, and displaced fishing effort concentrates around the border of the MPAs.

@@

Containing a carefully designed, well-enforced, and well-studied MPA network, the Channel Islands seems to be an ideal location to study the population-level effects of protected areas. High response ratios suggest that despite overall decreases in targeted biomass densities inside and outside MPAs relative to pre-MPA measurements (Fig.\@ref(fig:raw-density)), the MPAs may still be providing protection within their borders. But, these spatial response ratios are not necessarily an indicator of broader population-level effects. The targeted vs. non-targeted DiD strategy presents an alternative means of estimating the population-level effects of MPAs. While having its own strict caveats, this strategy provides some potential improvements over spatial response-ratios as a means of estimating population-level effects. While we estimate an uncertain but overall positive effect of the MPA network in its first few years, we are unable to detect a robust signal from 2012-2017. After 14 years of MPA protection we are left without a clear picture of the population-level effect of the Channel Island MPA network on biomass densities of targeted fin-fish species.

@@

# Supporting Information  {#si .unnumbered}

Detailed descriptions of the simulation model and DiD estimator (Appendix S1) are available online. The authors are solely responsible for the content and functionality of these materials. Queries (other than absence of the material) should be directed to the corresponding author.

@@

\newpage

@@

knitr::kable(equations,
             escape = FALSE,
             caption = "Variables for MPA simulations",
             format = "pipe",
             align = "l") %>% 
  kableExtra::kable_styling()

@@

\newpage

```{r coef-table, echo=FALSE}

coefs <- did_fit %>%
  broom.mixed::tidy(conf.int = TRUE, prob = 0.9) %>%
    rename(lower = conf.low,
         upper = conf.high) %>% 
  select(term, estimate, lower, upper) %>%
  rename(
    Mean = estimate,
    `5th Percentile` = lower,
    `95th Percentile` = upper
  )

coefs$term <- str_remove_all(coefs$term, "var_")

coefs$term <- str_replace_all(coefs$term, "targeted","Targeted by Fishing")

coefs$term <- str_replace_all(coefs$term, "year_bins","Year Block ")

coefs$term <- str_replace_all(coefs$term, ":","$\\\\times$")
